---
title: "2.2_Tumor_airway_comparisons_cutoffs_intersections_LUSC_CSFS_20250304"
output: html_document
date: "2025-03-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

- 2025/01/31: This version uses the LUSC data instead of LUAD data.

-2025/05/14: See the LUAD cutoffs intersections script for the same description:
Essentially, right now I have 2 candidate methods to filter to the genes of interest.
One is applying the cutoffs I defined in my progress report.
The other is using "sequential filtering" and going with the 75th percentile, then simple overlaps again.
I want to try both methods.
I know that the hypergeometric test shows that step 5 of my sequential filtering method has significant overlap for LUAD.
Now I want to revisit the progress report cutoffs method and see whether the overlap is significant by hypergeometric test.
If it is, I will probably go with that for the main part of the project, since we have all agreed that it is acceptable.
If not, I will make the argument for using the sequential filtering method because it revealed that when filtering for percentiles, there is very significant overlap.

# Comparing and filtering A1, TE, TM, and A2 to generate "linked genes" and "linked persistent genes" lists

## Reading in datasets to intersect
```{r}

# A1
GSE63127_CS_NS              <- read.table("../../2_Outputs/1_Airway_DEGs/GSE63127_CS_NS_limma_20250307.txt", header = TRUE)

# T-E
TCGA_LUSC_DEG               <- read.table("../../2_Outputs/2_Tumor_DEGs/TCGA_LUSC_limma_DEG_CSFS_20250307.txt")

# T-M
TCGA_LUSC_DMG               <- read.table("../../2_Outputs/3_Tumor_DMGs/TCGA_LUSC_limma_DMG_20250307.txt")

# A2
GSE7895_persistent_DEGs     <- read.table("../../2_Outputs/1_Airway_DEGs/GSE7895_persistent_DEGs_quantile_20250307.txt")

```


This is the method I was using in a sequential pipeline to try to test each assumption being made along the way, though I arbitrarily picked a 75th percentile log2FC cutoff.

## Using sequential filtering with 75th percentiles

```{r}
library(dplyr)

### First format them all the same way (this time not filtering for FDR)
A1_formatted <- GSE63127_CS_NS %>%
  dplyr::rename(., Gene = Gene.symbol, FDR_A1 = adj.P.Val, log2FC_A1 = logFC) %>%
  dplyr::select(., Gene, FDR_A1, log2FC_A1)

LUSC_TE_formatted<- TCGA_LUSC_DEG %>%
  dplyr::rename(., FDR_TE = FDR, log2FC_TE = log2FC, Gene = Gene) %>%
  dplyr::select(., Gene, FDR_TE, log2FC_TE)

LUSC_TM_formatted <- TCGA_LUSC_DMG  %>%
  dplyr::rename(., FDR_TM = FDR, log2FC_TM = log2FC) %>%
  dplyr::select(., Gene, FDR_TM, log2FC_TM)


print("Filter 1: Shared gene symbols between A1 and TE")
LUSC_A1_TE_shared_genes <- A1_formatted %>%
  dplyr::filter(Gene %in% intersect(LUSC_TE_formatted$Gene, A1_formatted$Gene))
nrow(LUSC_A1_TE_shared_genes)

LUSC_TE_A1_shared_genes <- LUSC_TE_formatted %>%
  dplyr::filter(Gene %in% intersect(LUSC_TE_formatted$Gene, A1_formatted$Gene))
nrow(LUSC_TE_A1_shared_genes)

# Note: The genes that end up being duplicates here are "DNAJC9-AS1", "POLR2J4", "RGS5", hence the difference in nrow for the two above, however I think these get lost in later steps so it does not really matter here.

print("Filter 2: FDR <= 0.05 individually for A1 and TE")
LUSC_A1_TE_sig <- LUSC_A1_TE_shared_genes %>%
  filter(FDR_A1 <= 0.05)
nrow(LUSC_A1_TE_sig)

LUSC_TE_A1_sig <- LUSC_TE_A1_shared_genes %>%
  filter(FDR_TE <= 0.05)
nrow(LUSC_TE_A1_sig)

print("Filter 3: FDR <= 0.05  in both A1 and TE")
LUSC_A1_TE_sig_all <- LUSC_A1_TE_sig %>%
  filter(Gene %in% LUSC_TE_A1_sig$Gene)
nrow(LUSC_A1_TE_sig_all)

LUSC_TE_A1_sig_all <- LUSC_TE_A1_sig %>%
  filter(Gene %in% LUSC_A1_TE_sig$Gene)
nrow(LUSC_A1_TE_sig_all)

print("Filter 4: 75th percentile |log2FC| in each")

## A1
# Calculate percentile cutoffs
top_cutoff_A1 <- quantile(LUSC_A1_TE_sig_all$log2FC_A1, 0.875, na.rm = TRUE)  # Top 12.5%
bottom_cutoff_A1 <- quantile(LUSC_A1_TE_sig_all$log2FC_A1, 0.125, na.rm = TRUE)  # Bottom 12.5%

# Filter genes in the top and bottom 25% of log2FC
LUSC_A1_TE_percentile <- LUSC_A1_TE_sig_all %>%
  filter(log2FC_A1 >= top_cutoff_A1 | log2FC_A1 <= bottom_cutoff_A1)
nrow(LUSC_A1_TE_percentile)

###

## TE
# Calculate percentile cutoffs
top_cutoff_TE <- quantile(LUSC_TE_A1_sig_all$log2FC_TE, 0.875, na.rm = TRUE)  # Top 12.5%
bottom_cutoff_TE <- quantile(LUSC_TE_A1_sig_all$log2FC_TE, 0.125, na.rm = TRUE)  # Bottom 12.5%

# Filter genes in the top and bottom 25% of log2FC
LUSC_TE_A1_percentile <- LUSC_TE_A1_sig_all %>%
  filter(log2FC_TE >= top_cutoff_TE | log2FC_TE <= bottom_cutoff_TE)
nrow(LUSC_TE_A1_percentile)

###

print("Filter 5: 75th percentile |log2FC| in both")
LUSC_A1_TE_merged_percentile <- LUSC_A1_TE_percentile %>%
  inner_join(., LUSC_TE_A1_percentile, by = "Gene")
nrow(LUSC_A1_TE_merged_percentile)

print("Filter 6: Same sign for log2FC")
LUSC_A1_TE_merged_percentile_sign <- LUSC_A1_TE_merged_percentile %>%
  filter(sign(log2FC_A1)==sign(log2FC_TE)) # Filter to genes with same signs
nrow(LUSC_A1_TE_merged_percentile_sign)

# Save this list of LUSC-A1-TE genes
#write.table(LUSC_A1_TE_merged_percentile_sign, "../2_Outputs/4_Linked_genes/LUSC_A1_TE_20250514.txt")

print("Merge with A2, same sign")
LUSC_A1_TE_A2 <- LUSC_A1_TE_merged_percentile_sign %>%
  inner_join(., GSE7895_persistent_DEGs, by = "Gene") %>%
  filter(sign(log2FC_A1)==sign(CS_NS_A2)) # Filter to genes with same sign
nrow(LUSC_A1_TE_A2)

# Save this list of LUSC-A1-TE-A2 genes
#write.table(LUSC_A1_TE_A2, "../2_Outputs/4_Linked_genes/LUSC_A1_TE_A2_20250514.txt")

```


## Hypergeometric tests for different filter steps
```{r}
print("Testing filter step 3")

N = nrow(LUSC_A1_TE_shared_genes) # All gene symbols shared between A1 and TE (filter step 1)
print(paste0("N = ",N))

n = nrow(LUSC_A1_TE_sig) # Number of genes with FDR <= 0.05 in A1 (filter step 3)
print(paste0("n = ",n))

m = nrow(LUSC_TE_A1_sig)  # Number of genes with FDR <= 0.05 in TE (filter step 3)
print(paste0("m = ",m))

k  = nrow(LUSC_A1_TE_sig_all) # Number of genes with FDR <= 0.05 in both A1 and TE
print(paste0("k = ",k))

p_value <- phyper(k - 1, m, N - m, n, lower.tail = FALSE)
print(paste0("p_value: ", p_value))



print("Testing last filter step (basically steps 5 and 6 are combined together)")

N = nrow(LUSC_A1_TE_sig_all) # The number of genes with FDR <= 0.05 in both A1 and TE (after filter step 3)
print(paste0("N = ",N))

n = nrow(LUSC_A1_TE_percentile) # Number of genes for top 75th percentile |log2FC| in each (here A1)
print(paste0("n = ",n))

m = nrow(LUSC_TE_A1_percentile)  # Number of genes for top 75th percentile |log2FC| in each (here TE)
print(paste0("m = ",m))

k  = nrow(LUSC_A1_TE_merged_percentile_sign) # Number of genes with top 75th percentile in both and with same sign
print(paste0("k = ",k))

p_value <- phyper(k - 1, m, N - m, n, lower.tail = FALSE)
print(paste0("p_value: ", p_value))

## Good, looks like the overlap is way more than expected by chance at this step, at least
```

I don't think that the hypergeometric test can apply to checking the A2 overlap step, because A1 and A2 cannot be called truly independent. Since they both measure genes with expression affected by smoking in airways (A1 in current smokers and A2 in former smokers), they will be expected to share many genes.

## Comparing and filtering A2 and "linked" genes to give "methylation related linked genes"

### Filtering by dataset intersection
```{r}
# First filter for FDR < 0.05
LUSC_TM_sig <- LUSC_TM_formatted %>%
  filter(FDR_TM <= 0.05)

# Then filter for 75th percentile
top_cutoff_TM <- quantile(LUSC_TM_sig$log2FC_TM, 0.875, na.rm = TRUE)  # Top 12.5%
bottom_cutoff_TM <- quantile(LUSC_TM_sig$log2FC_TM, 0.125, na.rm = TRUE)  # Bottom 12.5%

# Filter genes in the top and bottom 25% of log2FC
LUSC_TM_sig_percentile <- LUSC_TM_sig %>%
  filter(log2FC_TM >= top_cutoff_TM | log2FC_TM <= bottom_cutoff_TM)
nrow(LUSC_TM_sig_percentile)

LUSC_A1_TE_A2_TM <- LUSC_A1_TE_A2 %>%
  dplyr::inner_join(., LUSC_TM_sig_percentile, by = "Gene") %>%
  dplyr::filter(sign(log2FC_A1) != sign(log2FC_TM)) # Ensure signs are opposite
nrow(LUSC_A1_TE_A2_TM)

LUSC_A1_TE_A2_TM$Gene

## Save
#writeLines(LUSC_A1_TE_A2_TM$Gene, "../../2_Outputs/4_Linked_genes/LUSC_A1_TE_A2_TM_genes_20250528.txt")
#write.table(LUSC_A1_TE_A2_TM, "../../2_Outputs/4_Linked_genes/LUSC_A1_TE_A2_TM_20250610.txt")

```


